---
title: "Plot Model Projections Output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 2,
  fig.width = 8,
  fig.retina = 1,
  fig.align = "center"
)
```

The package `hubVis` contains an important function called 
`plot_step_ahead_forecasts()` and used to plot model projections output.

This function plot forecasts/scenario projections and optional truth data. 
Faceted plots could be created for multiple scenarios, locations, 
forecast dates, models, etc. 
Currently, the function can plot only quantiles data, with the possibility to 
add "median" information from the model projections.

For more information about the Hubverse standard format, please refer to the
[HubDocs website](https://hubdocs.readthedocs.io/en/latest/format/tasks.html).

The following vignette describes the principal usage of the 
`plot_step_ahead_forecasts()` function.

```{r lib}
library(hubVis)
```

All the plots are available in two output formats:

- "interactive" format: Plotly output object with interactive legend, hover
text, zoom-in and zoom-out options, etc.
- "static" format: ggplot2 output object
By default, the output plot is "interactive", but it can be change to "static"
by using the `interactive` parameter. See end of the document for examples.

## Load and Filter Data

The package contains 2 dataset that can be used for examples:

- `example_round1.csv`: example of model output for round associated with the 
origin date: "2021-03-07" (called "round 1"), target: "incident case", for the
US national level, from the
[example-complex-scenario-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-scenario-hub). The data set also contains an ensemble calculated by applying the 
function: `hubEnsembles::simple_ensemble(df_round1, agg_fun = "median")`

- `truth_data.csv`: example of target data from the [example-complex-scenario-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-scenario-hub). The data here comes from the `"target-data/US_inc_case.csv"` file.

### Load data

```{r load_data}
projection_path <- system.file("example_round1.csv", package = "hubVis")
projection_data <- read.csv(projection_path)
head(projection_data)

truth_path <- system.file("truth_data.csv", package = "hubVis")
truth_data <- read.csv(truth_path)
head(truth_data)
```

### Data Preparation

To fit the input format of the function, a `target_date` column needs to be 
calculated for the model projections data.

```{r data_prep}
projection_data <- dplyr::mutate(
  projection_data, target_date = as.Date(origin_date) + (horizon * 7) - 1)
head(projection_data)
```

## Plot

The function requires only 2 parameters:

- `forecast_data`: a `model_output_df` object, containing all the Hubverse
standard columns, and a `"target_date"` and a `"model_id"` column in the 
`hub_mdl_out_df format`. All forecasts in forecast_data will be plotted, 
i.e. all filtering needs to happen outside this function.
- ` truth_data`: a `data.frame` object containing the ground truth data, and 
with the columns: `time_idx` and `value`

### "Simple" plot

The `"projection_data"` and `"truth_data"` contain information for multiple 
location, and scenario.

To plot the model projections for the US, Scenario A:

```{r A_us}
# Pre-filtering
projection_data_A_us <- dplyr::filter(projection_data, 
                                      scenario_id == "A-2021-03-05", 
                                      location == "US")
# Class update
projection_data_A_us <- hubUtils::as_model_out_tbl(projection_data_A_us)
# Limit time_idx for layout reason
truth_data_us <- dplyr::filter(truth_data, location == "US", 
                               time_idx < min(projection_data$target_date) + 21,
                               time_idx > "2020-10-01")
```

```{r plot_A_us}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us)
```

By default, the 50%, 80% and 95% intervals are plotted, with a specific color 
per `model_id`.

It is possible to add a median line on the plot with the `use_median_as_point`
parameter:

```{r plot_A_us_med}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          use_median_as_point = TRUE)
```


### Facet plot

A "facet" (or subplot) plot can also be created for each scenario

```{r us}
# Pre-filtering
projection_data_us <- dplyr::filter(projection_data,
                                      location == "US")
# Class update
projection_data_us <- hubUtils::as_model_out_tbl(projection_data_us)
# Limit time_idx for layout reason
truth_data_us <- dplyr::filter(truth_data, location == "US", 
                               time_idx < min(projection_data$target_date) + 21,
                               time_idx > "2020-10-01")
```

```{r plot_us}
plot_step_ahead_forecasts(projection_data_us, truth_data_us, 
                          facet = "scenario_id")
```

The layout of the "facets" can be adjusted:

```{r plot_us_facet}
plot_step_ahead_forecasts(projection_data_us, truth_data_us, 
                          use_median_as_point = TRUE,
                          facet = "scenario_id", facet_scales = "free_x", 
                          facet_nrow = 2, facet_title = "bottom left")
```

A "facet" (or subplot) plot can also be created for each model. In this case, 
the legend will be adapted to return the `model_id` value.

```{r plot_facet_mod}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          facet = "model_id")
```

The legend can be removed with the parameter `show_legend = FALSE`.

```{r plot_facet_mod_noleg}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          facet = "model_id", show_legend = FALSE)
```

### Intervals

By default, the 50%, 80% and 95% intervals are plotted. However, it is possible 
to also plot the 90& intervals or a subset of these intervals.
When plotting 6 models or more, the plot will be reduced to show the highest 
intervals inputted (95% by default)

To illustrate this we will use the projections for only one model

```{r us_modl}
# Pre-filtering
projection_data_mod <- dplyr::filter(projection_data,
                                      location == "US", 
                                     model_id == "hub-ensemble")
# Class update
projection_data_mod <- hubUtils::as_model_out_tbl(projection_data_mod)
```


```{r plot_us_interv}
plot_step_ahead_forecasts(projection_data_mod, truth_data_us, 
                          use_median_as_point = TRUE, facet = "scenario_id", 
                          facet_nrow = 2, intervals = c(0.5, 0.8, 0.9, 0.95))
```

The opacity of the intervals can be adjusted:

```{r plot_us_interv_opa}
plot_step_ahead_forecasts(projection_data_mod, truth_data_us, 
                          use_median_as_point = TRUE, facet = "scenario_id", 
                          facet_nrow = 2, intervals = c(0.5, 0.8, 0.9, 0.95),
                          fill_transparency = 0.15)
```

No intervals can also be plotted:
```{r plot_us_no_interv}
plot_step_ahead_forecasts(projection_data_mod, truth_data_us, 
                          use_median_as_point = TRUE, facet = "scenario_id", 
                          facet_nrow = 2, intervals = NULL)
```

### Other parameters

Multiple other parameters are available to update the plot output. Here is some
examples of some parameters.

#### "Ensemble" layout

It is possible to assign a specific color and behavior to a specific `model_id`:
the associated value will be the top layer of the resulting plot.

```{r plot_us_ens}
plot_step_ahead_forecasts(projection_data_us, truth_data_us,
                          use_median_as_point = TRUE, 
                          facet = "scenario_id", facet_nrow = 2,
                          ens_name = "hub-ensemble", ens_color = "black")
```

#### Layout update

Multiple layout update are possible:

- Not showing the truth data in the plot:
```{r plot_us_td}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          plot_truth = FALSE)
```

- Change the top layer to the truth data:
```{r plot_us_top}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          top_layer = "truth")
```

- Add a title to the plot:
```{r plot_us_title}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          title = "Incident Case in the US")
```

- Change palette color and behavior:
    - The default palette can be changed. All the available palette names are
    available here:
```{r palette, fig.height = 6}
RColorBrewer::display.brewer.all()
```

```{r plot_us_palette}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          pal_color = "Dark2")
```
  
  - By default, separate colors will be used for each model. It is possible to use only blues for all models, by setting the `fill_by_model` parameter to `FALSE`
```{r plot_us_mod}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          fill_by_model = FALSE, use_median_as_point = TRUE)
```

- Interactive/Static plot:
```{r plot_stat}
plot_step_ahead_forecasts(projection_data_A_us, truth_data_us, 
                          interactive = FALSE)
```


For other parameter, please consult the documentation associated with 
the function:  
`?plot_step_ahead_forecasts`

