---
title: "Plot Model Projections Output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 2,
  fig.width = 8,
  fig.retina = 1,
  fig.align = "center"
)
```

The  `hubVis` package contains a function called 
`plot_step_ahead_model_output()` that can be used to plot model output that is 
in the format of forecasts or projects that look multiple horizons into the
future.

This function plots forecasts/scenario projections and optional truth data. 
Faceted plots can be created for multiple scenarios, locations, 
forecast dates, models, etc. 
Currently, the function can plot only quantile data, with the possibility to 
add "median" information from the model projections.

For more information about the Hubverse standard format, please refer to the
[HubDocs website](https://hubdocs.readthedocs.io/en/latest/format/tasks.html).

The following vignette describes the principal usage of the 
`plot_step_ahead_model_output()` function.

```{r lib}
library(hubVis)
library(hubUtils)
```

Plots are available in two output formats:

- "interactive" format: a [Plotly](https://plotly.com) output object with interactive legend, hover
text, zoom-in and zoom-out options, etc.
- "static" format: a [ggplot2](https://ggplot2.tidyverse.org/) output object.
By default, the output plot is "interactive", but it can be changed to "static"
by setting  the `interactive` parameter to FALSE. See end of the document for examples.

## Load and Filter Data

The package contains two datasets that will be used for the following examples:

- `example_round1.csv`: example of model output for a round associated with the 
origin date: "2021-03-07" (called "round 1"), target: "incident case", for the
US national level, from the
[example-complex-scenario-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-scenario-hub). The data set also contains an ensemble calculated by applying the 
function: `hubEnsembles::simple_ensemble(df_round1, agg_fun = "median")`

- `truth_data.csv`: example of target data from the [example-complex-scenario-hub](https://github.com/Infectious-Disease-Modeling-Hubs/example-complex-scenario-hub). The data here comes from the `"target-data/US_inc_case.csv"` file.

### Load data

```{r load_data}
projection_path <- system.file("example_round1.csv", package = "hubVis")
projection_data <- read.csv(projection_path)
projection_data <- as_model_out_tbl(projection_data)
head(projection_data)

truth_path <- system.file("truth_data.csv", package = "hubVis")
truth_data <- read.csv(truth_path)
head(truth_data)
```

### Data Preparation

The model output data in the `projection_data` object follows the structure 
of [the `model_out_tbl` class](https://infectious-disease-modeling-hubs.github.io/hubUtils/articles/connect_hub.html#structure-of-hubverse-datasets). 
This dataset is converted to a `model_out_tbl` object after being read-in above.
In addition to the standard requirements for this class, the plotting function 
in `hubVis` requires that the dataset have a `"target_date"` column, as this 
variable is used as the x-axis variable for the plot. This column does not 
exist in the raw example data, so we add it here:

```{r data_prep}
projection_data <- dplyr::mutate(
  projection_data, target_date = as.Date(origin_date) + (horizon * 7) - 1)
head(projection_data)
```

## Plot

The plotting function requires only 2 parameters:

- `model_output_data`: [a `model_out_tbl` object](https://infectious-disease-modeling-hubs.github.io/hubUtils/articles/connect_hub.html#structure-of-hubverse-datasets) containing all the Hubverse 
standard columns, including `"target_date"` and `"model_id"` columns. 
As all model_output in model_output_data will be plotted, any filtering needs to 
happen outside this function.

- ` truth_data`: a `data.frame` object containing the ground truth data, including 
the columns: `"time_idx"` and `"value"`.

### "Simple" plot

The `projection_data` and `truth_data` contain information for multiple 
locations, and scenarios.

To plot the model projections for the US, Scenario A:

```{r A_us}
# Pre-filtering
projection_data_A_us <- dplyr::filter(projection_data, 
                                      scenario_id == "A-2021-03-05", 
                                      location == "US")

# Limit time_idx for layout reason
truth_data_us <- dplyr::filter(truth_data, location == "US", 
                               time_idx < min(projection_data$target_date) + 21,
                               time_idx > "2020-10-01")
```

```{r plot_A_us}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us)
```

By default, the 50%, 80% and 95% intervals are plotted, with a specific color 
palette per `model_id`.

In general, it is hard to see multiple intervals when multiple models are 
plotted, so specifying only one interval can be useful:

```{r plot_A_us_one_int}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             intervals = 0.8)
```

It is also possible to add a median line on the plot with the `use_median_as_point`
parameter:

```{r plot_A_us_med}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             intervals = 0.8,
                             use_median_as_point = TRUE)
```

By default plots are interactive, but that can be easily switched to static:
```{r plot_A_us_med_static}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             intervals = 0.8,
                             use_median_as_point = TRUE, 
                             interactive = FALSE)
```

### Facet plot

A "facet" (or subplot) plot can also be created for each scenario

```{r us}
# Pre-filtering
projection_data_us <- dplyr::filter(projection_data,
                                    location == "US")
# Limit time_idx for layout reason
truth_data_us <- dplyr::filter(truth_data, location == "US", 
                               time_idx < min(projection_data$target_date) + 21,
                               time_idx > "2020-10-01")
```

```{r plot_us}
plot_step_ahead_model_output(projection_data_us, truth_data_us, 
                             facet = "scenario_id")
```

The layout of the "facets" can be adjusted:

```{r plot_us_facet}
plot_step_ahead_model_output(projection_data_us, truth_data_us, 
                             use_median_as_point = TRUE,
                             facet = "scenario_id", facet_scales = "free_x", 
                             facet_nrow = 2, facet_title = "bottom left")
```

A "facet" (or subplot) plot can also be created for each model. In this case, 
the legend will be adapted to return the `model_id` value.

```{r plot_facet_mod}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             facet = "model_id")
```

The legend can be removed with the parameter `show_legend = FALSE`.

```{r plot_facet_mod_noleg}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             facet = "model_id", show_legend = FALSE)
```


### Intervals

By default, the 50%, 80% and 95% intervals are plotted. However, it is possible 
to also plot the 90% intervals or a subset of these intervals.
When plotting 6 or more  models, the plot will be reduced to show the widest 
intervals provided (95% by default).

To illustrate this we will use the projections for only one model

```{r us_modl}
# Pre-filtering
projection_data_mod <- dplyr::filter(projection_data,
                                     location == "US", 
                                     model_id == "hub-ensemble")
```


```{r plot_us_interv}
plot_step_ahead_model_output(projection_data_mod, truth_data_us, 
                             use_median_as_point = TRUE, facet = "scenario_id", 
                             facet_nrow = 2, intervals = c(0.5, 0.8, 0.9, 0.95))
```


The opacity of the intervals can be adjusted:

```{r plot_us_interv_opa}
plot_step_ahead_model_output(projection_data_mod, truth_data_us, 
                             use_median_as_point = TRUE, facet = "scenario_id", 
                             facet_nrow = 2, intervals = c(0.5, 0.8, 0.9, 0.95),
                             fill_transparency = 0.15)
```

Plots without intervals are also possible:
```{r plot_us_no_interv}
plot_step_ahead_model_output(projection_data_mod, truth_data_us, 
                             use_median_as_point = TRUE, facet = "scenario_id", 
                             facet_nrow = 2, intervals = NULL)
```

### Other parameters

Several other parameters are available to update the plot output. Here is some
examples of some parameters.

#### "Ensemble" layout

It is possible to assign a specific color and behavior to a specific `model_id`.
Typically, this is done to highlight an ensemble, so the name for these 
arguments are `ens_name` and `end_color`. The model specified by `ens_name` 
will be the top layer of the resulting plot.

```{r plot_us_ens}
plot_step_ahead_model_output(projection_data_us, truth_data_us,
                             use_median_as_point = TRUE, 
                             facet = "scenario_id", facet_nrow = 2,
                             ens_name = "hub-ensemble", ens_color = "black",
                             intervals = 0.8)
```

#### Layout update

Multiple layout update are possible:

- Not showing the truth data in the plot:
```{r plot_us_td}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             plot_truth = FALSE)
```

- Change the top layer to the truth data:
```{r plot_us_top}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             top_layer = "truth")
```

- Add a title to the plot:
```{r plot_us_title}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             title = "Incident Cases in the US")
```

- Change palette color and behavior:
    - The default palette can be changed. All the available palette names are
    available here:
```{r palette, fig.height = 7.5}
RColorBrewer::display.brewer.all()
```

```{r plot_us_palette}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             pal_color = "Dark2")
```
  
  - By default, separate colors will be used for each model. 
  
The `fill_by` parameter can be change to another valid column names to change
the legend and colors attributes to this new column.
```{r plot_facet_mod_scenleg}
plot_step_ahead_model_output(projection_data_us, truth_data_us, 
                             facet = "model_id", fill_by = "scenario_id")
```

  
It is possible to use only blues for all models, by setting the `pal_color` parameter to `NULL`. This might be especially useful when used for many models in conjunction with highlighting the ensemble forecast using the `ens_name` and `ens_color` argument.
```{r plot_us_mod}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             intervals = 0.8,
                             ens_name = "hub-ensemble", ens_color = "black",
                             pal_color = NULL, use_median_as_point = TRUE)
```

- Interactive/Static plot:
```{r plot_stat}
plot_step_ahead_model_output(projection_data_A_us, truth_data_us, 
                             interactive = FALSE)
```


For other parameters, please consult the documentation associated with 
the function:  
`?plot_step_ahead_model_output`

